Precise spiking motifs in neurobiological and neuromorphic data

### 2022-10-17 manuscript.tex
4.3
abstract

% Authors, for the paper (add full first names)
\Author{
Antoine Grimaldi $^{1,\dagger,\ddagger}$\orcidA{}, 
Amélie Gruel $^{2,\dagger,\ddagger}$, 
Jean Martinet $^{2,}$ 
and Laurent Perrinet $^{1,}$\orcidD{}*}

@@

% Abstract (Do not insert blank lines, i.e. \\) 
\abstract{
Why do neurons communicate through spikes? By definition, a spike, or action potential, is a all-or-none event ---it can occur or not without further details--- at asynchronous timings, i.e.~it can occur at any continuous time - differentially to the discretized timing classically used in digital processing. In the living world, neural systems almost systematically use this so-called event-based representation. A better understanding of this phenomenon remains a fundamental challenge in neurobiology in order to better interpret the masses of recorded data. It is also an emerging challenge in computer science to allow the efficient exploitation of a new class of sensors and event-based computers, called neuromorphic, which could allow significant gains in computing time and energy consumption ---a major societal challenge in the age of the digital economy and of global warming.
}

@@

%There are more applications to image processing using sparse spiking representations. Using the core computational unit defined, extension of the computation to a topographic representation similar to that observed in the primary visual cortex of mammals. design of micro-circuits with specific lateral interactions will allow us to design efficient micro-circuits for the sparse representation of images.
A predictive model defined in the framework of the free-energy principle was able to account for temporal delays in the nervous system, both at the sensory and motor levels~\citep{perrinet_active_2014}.  %
%and behavioral~\citep{khoei_flash-lag_2017} levels. Extending this knowledge to the optimization of delays in a SNN will provide a breakthrough in the efficiency of these networks.
Such a model was implemented at the network level by showing that motion-based prediction explains the role of tracking in motion extrapolation interpolation~\citep{khoei_motion-based_2013}. Note that this motion-based prediction is implemented in the weights of the neural network by defining connections between neurons with specific delays.  Two neurons which were selective to specific motions were connected if the delay was coherent with the change in the position of their respective receptive fields~\citep{hogendoorn_predictive_2019}. This was aso implemented in a spiking neural network and reproduced the observation that neural activty was maintained during the trajectory of a smoothly moving dot even if it was momentarily blanked~\citep{kaplan_anisotropic_2013}. This led in particular to the proposal that such delay-based computations could explain diverse perceptual mechanisms, such as the flash-lag illusion~\citep{khoei_flash-lag_2017}. 
We will review some more applications of using precise spiking motifs in neuromorphic engineering, but will first review current evidence in neurobiology, that is, from actual recordings in biological tissues.
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \section{Precise spiking motifs in neurobiology}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
% In Bellec~\citep{bellec_fitting_2021}, authors fit summary statistics of neural data with a differentiable spiking network simulator. The loss function is the cross entropy (following Bernouilli hypothesis with a GLM where each unit is modelled with a SRM neuron~\citep{gerstner_time_1995} with recurrent dynamics). The model uses samples and measure method sto include latent / hidden neurons. In particular, it comes with  \href{code}{https://github.com/EPFL-LCN/pub-bellec-wang-2021-sample-and-measure}. It uses the publicly available \href{V1-dataset}{http://crcns.org/data-sets/vc/pvc-11}, which allows to supervise the model with the input being the movie and the output the spikes recorded.

 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Decoding neural activity using spike distances}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
In most generic computational models the neural activity is assumed to be encoded in the firing rate. For instance, linear non linear L-NL models, the output is assumed to be distributed according to a Poisson law. As such a simple decoding strategy is to asscume it is to b inferned for a given tuning curves (Jayazeri) or simply by a simple regression~\citep{berens_fast_2012}. This latter model assumes a Bernoulli model for the generation of spikes such that the decoding amounts to a single-layer logistic regression.

>>>>

% Authors, for the paper (add full first names)
\Author{
Antoine Grimaldi $^{1}$% $^{1,\dagger,\ddagger}$
\orcidA{}, 
Amélie Gruel $^{2}$, % $^{2,\dagger,\ddagger}$, 
Jean Martinet $^{2}$,  
and Laurent Perrinet $^{1,}$\orcidD{},*}

@@

% Abstract (Do not insert blank lines, i.e. \\) 
\abstract{
Why do neurons communicate through spikes? By definition, a spike, or action potential, is a all-or-none event ---it can occur or not without further details--- at timings asynchronous within a cell assembly, i.e.~it can occur at any continuous time. This stands in stark contrast to the discretized timing classically used in digital processing at the base of modern-day neural networks. In the living world, neural systems almost systematically use this so-called event-based representation. A better understanding of this phenomenon remains a fundamental challenge in neurobiology in order to better interpret the masses of recorded data. It is also an emerging topic in computer science to allow the efficient exploitation of a new class of sensors and event-based computers, called neuromorphic, which could allow significant gains in computing time and energy consumption ---a major societal mission in the age of the digital economy and of global warming.
}

@@

%There are more applications to image processing using sparse spiking representations. Using the core computational unit defined, extension of the computation to a topographic representation similar to that observed in the primary visual cortex of mammals. design of micro-circuits with specific lateral interactions will allow us to design efficient micro-circuits for the sparse representation of images.
A predictive model defined in the framework of the free-energy principle was able to account for temporal delays in the nervous system, both at the sensory and motor levels~\citep{perrinet_active_2014}.  %
%and behavioral~\citep{khoei_flash-lag_2017} levels. Extending this knowledge to the optimization of delays in a SNN will provide a breakthrough in the efficiency of these networks.
Such a model was implemented at the network level by showing that motion-based prediction explains the role of tracking in motion extrapolation interpolation~\citep{khoei_motion-based_2013}. Note that this motion-based prediction is implemented in the weights of the neural network by defining connections between neurons with specific delays.  Two neurons which were selective to specific motions were connected if the delay was coherent with the change in the position of their respective receptive fields~\citep{hogendoorn_predictive_2019}. This was aso implemented in a spiking neural network and reproduced the observation that neural activty was maintained during the trajectory of a smoothly moving dot even if it was momentarily blanked~\citep{kaplan_anisotropic_2013}. This led in particular to the proposal that such delay-based computations could explain diverse perceptual mechanisms, such as the flash-lag illusion~\citep{khoei_flash-lag_2017}. 
We will review some more applications of using precise spiking motifs in neuromorphic engineering, but will first review current evidence in neurobiology, that is, from actual recordings in biological tissues.
 \section{Precise spiking motifs in neurobiology}
% In Bellec~\citep{bellec_fitting_2021}, authors fit summary statistics of neural data with a differentiable spiking network simulator. The loss function is the cross entropy (following Bernouilli hypothesis with a GLM where each unit is modelled with a SRM neuron~\citep{gerstner_time_1995} with recurrent dynamics). The model uses samples and measure method sto include latent / hidden neurons. In particular, it comes with  \href{code}{https://github.com/EPFL-LCN/pub-bellec-wang-2021-sample-and-measure}. It uses the publicly available \href{V1-dataset}{http://crcns.org/data-sets/vc/pvc-11}, which allows to supervise the model with the input being the movie and the output the spikes recorded.
In most generic computational models the neural activity is assumed to be encoded in the firing rate. For instance, linear non linear L-NL models, the output is assumed to be distributed according to a Poisson law. As such a simple decoding strategy is to asscume it is to b inferned for a given tuning curves (Jayazeri) or simply by a simple regression~\citep{berens_fast_2012}. This latter model assumes a Bernoulli model for the generation of spikes such that the decoding amounts to a single-layer logistic regression. As such, all these models are dependent on  a core definition of spike measures and we will review here how precise spiking motifs are taken into account in various neurobiological results.
\subsection{Decoding neural activity using spike distances}
